\chapter{DAG Construction Formalism}\label{app:dag-formalism}

This appendix gives the full architectural and mathematical detail
behind the taxonomy construction pipeline summarised in
Section~\ref{sec:arch-taxonomy} (Chapter~\ref{ch:system-architecture})
and Section~\ref{sec:meth-construction-summary}
(Chapter~\ref{ch:methodology}). The DAG construction pipeline is a
necessary prerequisite that the arena evaluation runs on, not the
thesis's own headline contribution; this appendix exists so the exact
mechanics are preserved and traceable without requiring the reader to
work through them before reaching the arena's methodology and
results.

\section{Construction Pipeline Walkthrough}\label{sec:app-dag-pipeline}

\subsection{Seed Initialisation from Ground-Truth Domains}

Before iterative refinement begins, each corpus query is assigned to
the anchor node corresponding to its MMLU-Pro ground-truth domain label.
This creates 14 depth-0 partitions---one per MMLU-Pro domain---each
instantiated as a \texttt{GraphNode} with \texttt{isAnchor = true},
\texttt{depth = 0}, and an initial member set $Q_D^{\text{canonical}}$
equal to the set of queries carrying label $D$.

The anchor flag enforces three invariants at the data-model level.
First, anchor nodes are exempt from all destructive topology
operations: \texttt{TaxonomyMerger} raises a
\texttt{TopologyViolationException} if any merge, prune, or collapse
operation targets an anchor.
Second, anchor nodes are never removed from the DAG; they appear in
every serialised snapshot with the same UUID and depth.
Third, the anchor's \texttt{sliceDim} is fixed at $d = 256$, the
single embedding dimension used at every depth
(Section~\ref{subsec:dim-schedule}).

The ground-truth label assignment serves as initialisation only.
The trickle routing phase may reassign a query to a geometrically more
appropriate anchor, so that the adapted partition
$Q_D^{\text{adapted}}$ at the end of construction need not equal
$Q_D^{\text{canonical}}$.
The distinction is recorded explicitly in the snapshot and is
material for the validation comparison: the comparison is performed
using $Q_D^{\text{adapted}}$, which reflects the geometry-corrected
membership rather than the editorial label.

\subsection{Iterative Six-Phase Refinement Loop}

After anchor initialisation, \texttt{TaxonomyEngine} executes a
six-phase loop over the current DAG state.
Each iteration processes all non-converged nodes; the loop runs for
a fixed iteration budget, with completion semantics defined in
Section~\ref{subsec:quiescence}.
Table~\ref{tab:pipeline} summarises the phases.

\begin{table}[h]
\centering
\caption{Taxonomy construction phases, implementing classes, and outputs.}
\label{tab:pipeline}
\begin{tabular}{clll}
\toprule
Phase & Responsibility & Implementing class & Output \\
\midrule
1 & MRL prefix selection and $\ell_2$-normalisation
& \texttt{EmbeddingService}
& Normalised slice matrix $X \in \mathbb{R}^{N \times 256}$ \\
2 & vMF MLE and NiW posterior fitting
& \texttt{TaxonomyFitter}
& $(\hat{\boldsymbol{\mu}}, \hat{\kappa})$ per node;
NiW hyperparameters $(m_N, \kappa_N, \nu_N, \Lambda_N)$ \\
3 & Soft routing (skipped in iteration 1)
& \texttt{TaxonomyTrickler}
& Updated weighted member sets; residual hits \\
4 & Adaptive splitting
& \texttt{TaxonomySplitter}
& New child nodes where the separation score $s \geq \epsilon_{\text{sep}}$ \\
5 & Topological optimisation
& \texttt{TaxonomyMerger}
& Merged siblings; pruned starved leaves;
cross-links; transitive reduction \\
6 & Convergence check
& \texttt{TaxonomyEngine}
& GED quiescence streak; early stop or next iteration \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Phase 1 --- Embedding selection.}
At every depth, \texttt{EmbeddingService} extracts the fixed
256-dimensional prefix slice from the pre-computed MRL vectors and
$\ell_2$-normalises each row to the unit hypersphere.

\textbf{Phase 2 --- vMF/NiW fitting.}
\texttt{TaxonomyFitter} fits a single-component vMF to each node's
member embeddings, producing the directional centroid
$(\hat{\boldsymbol{\mu}}, \hat{\kappa})$, and subsequently a
Normal-inverse-Wishart posterior used for Bayesian membership scoring
in Phase~3.
Statistical detail is given in Section~\ref{sec:meth-vmf}.

\textbf{Phase 3 --- Trickle routing.}
\texttt{TaxonomyTrickler} routes each query down the DAG by scoring
each child with its own vMF log-density and taking a temperature-1
softmax over the siblings; every child whose posterior responsibility
clears the \texttt{membershipFloor} is admitted as a weighted
membership destination (Section~\ref{sec:meth-routing}).
Positions where no child clears the floor are recorded as residual
hits.
This phase is skipped entirely in iteration~1: the ground-truth
bootstrap assignment is kept for the first split/optimise/refit pass,
and geometric reassignment starts at iteration~2
(Section~\ref{subsec:migration}).
DAG polyhierarchy arises from Phase~5 topology operations, not from
construction-time routing.

\textbf{Phase 4 --- Adaptive splitting.}
\texttt{TaxonomySplitter} proposes partitions of non-anchor
nodes via vMF-EM clustering in a PCA-projected subspace, accepting only
splits whose chance-corrected separation score
(Section~\ref{sec:meth-split}) clears the threshold
$\epsilon_{\text{sep}}$ and all of whose children satisfy a minimum
size constraint $n_{\min}$ (values in
Chapter~\ref{ch:experimental-design}).

\textbf{Phase 5 --- Topological optimisation.}
\texttt{TaxonomyMerger} executes a sequence of anchor-exempt
operations: merge of similar siblings (chance-corrected separation
below $\epsilon_{\text{sep}}$); pruning of starved leaves with queries
reassigned to the nearest sibling; removal of structurally redundant
single-child nodes (Section~\ref{sec:meth-topology});
soft-membership refitting via \texttt{refitMuPerIteration}, subject to
a no-update band that freezes a node's centroid once its refit
direction agrees with the previous one to within cosine $0.9975$; and
bridge
insertion from two distinct sources, followed by transitive reduction
to maintain the minimal DAG.
Bridge insertion replaces a single global cross-link threshold with
two gated sources described in
Section~\ref{subsec:polyhierarchy}: \textit{Source-A} residual-node
routing and \textit{Source-B} cross-domain bridging.

\textbf{Phase 6 --- Convergence check.}
At the end of each iteration, \texttt{TaxonomyEngine} computes the
graph edit distance to the preceding iteration's DAG and either stops
construction early once the structure has been quiescent for five
consecutive iterations, triggers the next iteration, or, when the
budget is exhausted, serialises the DAG as an immutable snapshot
(Section~\ref{subsec:quiescence}).

\subsection{Cross-Domain Migration}\label{subsec:migration}\label{sec:meth-migration}

At bootstrap, each anchor $D_k$ is initialised with the canonical
partition---every query placed on its ground-truth anchor with weight
$1.0$---and a vMF/NiW self-model fit at $d = 256$; the anchor's
identity, depth, and \texttt{isAnchor} flag are set once and never
modified.

This bootstrap assignment is deliberately preserved through the first
construction iteration: Phase~3 trickle reassignment is skipped in
iteration~1, so the first split/optimise/refit pass operates on the
pure ground-truth partition, and geometric reassignment starts at
iteration~2, once real sub-domain structure exists for routing to act
on.
(Re-routing immediately in iteration~1---even with a ground-truth
bias nudge, \texttt{enableGtWarmStart}, which is disabled---let
queries drift onto merely-similar wrong anchors before any structure
existed; the iteration-1 skip removes the need for any such bias.)
From iteration~2 onward, queries are reassigned by the standard
membership-floor trickle routing of Section~\ref{sec:meth-routing};
\textit{migration} is emergent from this re-routing rather than a
separately gated mechanism with its own margin, and the
\textit{adapted partition} $Q_{D_k}^{\text{adapted}}$ is defined as
the anchor-level assignment after construction completes.
Anchor nodes are exempt from all destructive topology operations
regardless of migration volume, enforced by a
\texttt{TopologyViolationException} at the data-model level.

Three downstream effects follow from migration.
First, the vMF self-model of each anchor is refit in Phase~2 of
subsequent iterations on $Q_{D_k}^{\text{adapted}}$, not
$Q_{D_k}^{\text{canonical}}$.
Second, all sub-domain structure induced below anchor $D_k$ is built
from $Q_{D_k}^{\text{adapted}}$, meaning that a query that migrated
into $D_k$ may reach a leaf that is thematically distant from its
MMLU-Pro label.
Third, the validation comparison (Section~\ref{sec:meth-validation})
is performed using $Q_{D_k}^{\text{adapted}}$ at the anchor level.
The migration rate
$m_{D_k} = |Q_{D_k}^{\text{adapted}} \setminus
Q_{D_k}^{\text{canonical}}| / |Q_{D_k}^{\text{canonical}}|$ is
reported as a diagnostic descriptor: it quantifies the fraction of
queries that changed anchor assignment and contextualises the
interpretation of the validation comparison.
The primary RQ2 metric is the fidelity difference
$\Delta\tau = \tau_{\text{adapted}} - \tau_{\text{canonical}}$---the
Kendall rank-agreement gain from using geometry-corrected rather than
editorial-label partitions when computing anchor-level accuracy---with
the Spearman-based $\Delta\rho$ reported secondarily.

\subsection{Completion and Quiescence Semantics}\label{subsec:quiescence}

Construction termination is criterion-based with a budget ceiling:
the loop runs for at most \texttt{numIterations} iterations (value
specified in Chapter~\ref{ch:experimental-design}), and a structural
early-stopping gate---enabled in the canonical configuration
(\texttt{enableEarlyStopping})---can terminate it earlier.
At the end of each iteration, the \texttt{TaxonomyStabilizer} computes
the graph edit distance (GED) between the current DAG
$\mathcal{G}^{(t)}$ and the preceding iteration's
$\mathcal{G}^{(t-1)}$, counting node additions and deletions plus
parent--child edge additions and deletions, and normalises it by the
current edge count.
After a minimum iteration count (proportional to the number of
domains, floored at $5$), an iteration counts as quiescent when this
relative GED falls at or below the quiescence threshold
(\texttt{gedThreshold}, $0.005$); construction stops early once five
consecutive iterations are quiescent.
A leaf-level log-semantic-volume trajectory is additionally tracked as
a convergence diagnostic.
If the streak never completes, the budget is exhausted and the
snapshot is frozen as-is; downstream analyses can distinguish
quiescent from still-evolving taxonomies from the recorded GED
trajectory.

\section{Directional Node Model}\label{sec:meth-vmf}

\subsection{von Mises--Fisher Density and MLE}

The vMF distribution on $\mathcal{S}^{d-1}$ with mean direction
$\boldsymbol{\mu}$ and concentration $\kappa \geq 0$ has density
\[
p(\mathbf{x} \mid \boldsymbol{\mu}, \kappa)
= C_d(\kappa) \exp\!\left(\kappa \boldsymbol{\mu}^\top \mathbf{x}\right),
\qquad
C_d(\kappa) = \frac{\kappa^{d/2 - 1}}{(2\pi)^{d/2}\, I_{d/2 - 1}(\kappa)}.
\]
Given $N$ unit-vector observations, the MLE of $\boldsymbol{\mu}$ is
the normalised sample mean $\hat{\boldsymbol{\mu}} =
\bar{\mathbf{x}} / \|\bar{\mathbf{x}}\|_2$.
The Banerjee closed-form approximation~\parencite{banerjee2005vmf}
\[
\hat{\kappa}_{0}
= \frac{\bar{R}\, d - \bar{R}^3}{1 - \bar{R}^2},
\qquad \bar{R} = \|\bar{\mathbf{x}}\|_2,
\]
avoids iterative root-finding for $\kappa$ and is accurate for
$d \geq 4$, $N \geq 5$, but is only an approximation to the true MLE
and is biased upward in high dimensions at
small sample sizes---an $O(d/N)$ bias that becomes severe when the
ratio $d/N$ is large.

Three refinements are applied, all reflecting what the
implementation actually executes.
First, the closed-form value seeds up to five Newton--Raphson steps
on the exact likelihood equation $A_d(\kappa) = \bar{R}$, where
$A_d(\kappa) = I_{d/2}(\kappa)/I_{d/2-1}(\kappa)$, using
$A_d'(\kappa) = 1 - A_d(\kappa)^2 - \tfrac{d-1}{\kappa}A_d(\kappa)$;
the seed is close enough that convergence typically occurs within one
or two steps, yielding $\hat{\kappa}_{\text{ML}}$.
Second, the refined estimate is multiplied by a small-sample
shrinkage factor
\[
\hat{\kappa}
= \hat{\kappa}_{\text{ML}} \cdot \frac{N - 1}{N + d - 2},
\]
a multiplicative correction in the spirit of the high-accuracy
$\kappa$ estimators of Hornik and Gr\"{u}n~\parencite{hornik2014vmf},
which pulls the estimate down toward its small-sample
expectation.
Third, the shrunk estimate is blended with a parent-anchored
empirical-Bayes prior $\kappa_{\text{parent}}$ (the mean of the
node's parents' concentrations, or a default prior at the root):
\[
\kappa_{\text{final}}
= (1 - \alpha(\rho))\,\hat{\kappa}
+ \alpha(\rho)\,\kappa_{\text{parent}},
\qquad
\alpha(\rho) =
\begin{cases}
0 & \rho \leq 2,\\
(\rho - 2)/8 & 2 < \rho \leq 10,\\
1 & \rho > 10,
\end{cases}
\]
where $\rho = d/n$ is the dimension-to-effective-sample-size ratio,
so a node's own data dominate when it is well supported and the
parent prior takes over smoothly as support thins.
The joint choice of fixed dimension and minimum leaf size should be
read against this ramp honestly: at $d = 256$ and $n_{\min} = 50$
(Section~\ref{sec:exp-env}), a minimum-size leaf sits at
$\rho \approx 5.1$ and hence at a blend weight of
$\alpha \approx 0.39$---partially prior-anchored, not comfortably
outside any special regime; only nodes with $n \gtrsim 128$ are
estimated from their own data alone, and the fully prior-dominated
$\rho > 10$ regime is additionally logged with a warning.
Numerically stable evaluation of $\log I_\nu(\kappa)$ via the Debye
asymptotic expansion is given in Appendix~\ref{app:numerics}.

\subsection{NiW Auxiliary Regulariser}

A Normal-inverse-Wishart (NiW) posterior is fit to each node's member
embeddings as an auxiliary Bayesian summary of the node's local
geometry; it no longer participates in routing, which is vMF-based
throughout (Section~\ref{sec:meth-routing}).
The NiW is \textit{not} the conjugate prior for the vMF---the vMF's
conjugate is the vMF itself~\parencite{gopal2014mises}---but is used
as an engineering regulariser whose posterior predictive is a
multivariate Student-$t$, accurate when $\kappa$ is moderate to large
so that the tangent-plane approximation to $\mathcal{S}^{d-1}$ holds.
At near-uniform nodes ($\kappa < 1$) the approximation degrades; such
nodes are flagged as geometric leaves and excluded from splitting.

The prior mean is the parent's vMF centroid, the prior pseudocount is
$\kappa_0 = 1.0$, the prior degrees of freedom are
$\nu_0 = d + 2$, and the prior scale is
$\boldsymbol{\Lambda}_0 = \lambda_0 \mathbf{I}_{d}$ with
$\lambda_0 = 1/(\hat{\kappa}_{\text{parent}} d)$.
Because the construction fixes $d = 256$ at every depth
(Section~\ref{sec:meth-mrl}), a diagonal approximation to
$\boldsymbol{\Lambda}_N$ is stored at all depths, reducing memory from
$O(d^2)$ to $O(d)$ per node and making cross-depth node comparisons
directly valid without down-projection.
The posterior update equations and the resulting Student-$t$
predictive density are given in Appendix~\ref{app:numerics}; in the
current implementation the stored posterior serves diagnostic roles
(notably the leaf log-semantic-volume convergence diagnostic of
Section~\ref{subsec:quiescence}), while all routing, migration, and
bridge-admission decisions operate on the vMF responsibilities of
Section~\ref{sec:meth-routing}.
Node centroids are refit directly to the fresh weighted MLE each
iteration; an earlier exponential-moving-average smoothing of the
centroid was removed after matched A/B runs showed it amplifies
oscillation rather than damping it.

\section{Fixed-Dimension Matryoshka Embedding Slice}\label{sec:meth-mrl}

MRL~\parencite{kusupati2022mrl} trains a single embedding model such
that every prefix slice $\mathbf{x}_{1:m}$ is itself a meaningful
embedding, allowing the full vector to be stored once and sliced at
query time.
Its key property is \textit{coarse-to-fine information packing}: the
most discriminative semantic directions are concentrated in the
leading coordinates, so any prefix $\mathbf{x}_{1:m'}$ is a coherent
lower-dimensional representation rather than a truncated artefact.

TaxoArena uses a single fixed slice
\[
d(\ell) = 256 \quad \text{for all depths } \ell,
\]
rather than a per-depth dimension schedule.
The choice balances two pressures.
A larger $d$ increases angular resolution and improves the
discriminability of fine-grained sub-topic splits, but the vMF
concentration estimate degrades as $d/N$ grows
(Section~\ref{sec:meth-vmf}); at $d = 256$ the fully prior-dominated
regime ($d/N > 10$) is crossed only below $N \approx
26$, and the $n_{\min} = 50$ floor keeps every fitted leaf above
that---though a minimum-size leaf still sits on the empirical-Bayes
blend ramp ($\alpha \approx 0.39$), i.e.\ partially prior-anchored
(Section~\ref{sec:meth-vmf}).
Fixing a single dimension additionally means all nodes---across all
depths---live on the same unit hypersphere, so sibling-merge,
passthrough-collapse, and bridge-coherence comparisons require no
cross-depth down-projection.
The coarse-to-fine property of MRL is exploited not by widening the
slice with depth but by the recursive structure itself: shallow nodes
separate coarse domain-level directions, while deeper splits resolve
progressively finer sub-topic structure within the same 256-dimensional
geometry.
(One qualification: split \emph{proposals} are clustered in a
PCA-projected subspace of 32, 64, or 128 dimensions depending on
cluster size, Section~\ref{sec:meth-split}; all fitting, routing, and
acceptance decisions remain at the fixed 256 dimensions.)
Re-normalisation after slicing is mandatory, since MRL guarantees
semantic meaningfulness of $\mathbf{x}_{1:d}$ but not unit norm:
\(
\mathbf{x}^{(\ell)}
= \mathbf{x}_{1:d} / \|\mathbf{x}_{1:d}\|_2.
\)
The dimension itself is ablated (A1, Section~\ref{sec:appendix-ablations})
at $\{128, 256, 512\}$ to confirm that $d = 256$ is a
quality-stable operating point rather than a fragile one.

\section{Fixed Embedding Dimension}\label{subsec:dim-schedule}

The effective embedding dimension is fixed at $d = 256$ for every
recursion depth: the coarse-to-fine progression from domain-level
separation to fine sub-topic structure is carried by the recursion
itself, not by a per-depth dimension schedule, exploiting the
multi-scale property of MRL embeddings~\parencite{kusupati2022mrl}
reviewed in Chapter~\ref{ch:literature-review}.
The fixed dimension places every node, at every depth, on the same
unit hypersphere, so sibling-merge, passthrough-collapse, and
bridge-coherence comparisons are directly valid without cross-depth
projection; its coherence with the minimum-cluster-size floor, which
keeps every fitted leaf outside the regime where the vMF concentration
estimate becomes unreliable, is established in
Section~\ref{sec:meth-vmf}.

\section{Adaptive Splitting Oracle}\label{sec:meth-split}

Split proposals are generated by vMF-EM clustering with a spherical
$k$-means$++$-style initialisation: the first centroid is the node's
normalised sample centroid and subsequent centroids are chosen by
maximin cosine distance to the existing centres, avoiding the
dual-collapse failure mode of random initialisation in high
dimensions.
For efficiency, the clustering itself runs in a PCA-projected
subspace whose dimension depends on cluster size (32 dimensions below
100 members, 64 below 500, 128 otherwise); the resulting partition is
then re-fit and evaluated in the fixed 256-dimensional routing
geometry, so no acceptance decision is made in the reduced space.

A proposed partition $\{S_c\}$ of a node's members is accepted or
rejected by a \emph{chance-corrected separation score}.
For a set $S$ of unit vectors, define the total pairwise cosine
dissimilarity
\[
W(S) = |S|^2 - \Big\|\sum_{\mathbf{x} \in S} \mathbf{x}\Big\|_2^2
= \sum_{i \neq j \in S} \left(1 - \mathbf{x}_i^\top \mathbf{x}_j\right),
\]
computable in $O(|S|\,d)$ from the member sum alone.
The observed within-cluster scatter of the partition is
$\sum_c W(S_c)$; under a uniformly random partition of the same $n$
points into the same cluster sizes $\{n_c\}$, its exact expectation is
\[
\mathbb{E}\Big[\sum_c W(S_c)\Big]
= W(S_{\text{total}}) \cdot
\sum_c \frac{n_c (n_c - 1)}{n (n - 1)},
\]
because each unordered pair lands inside cluster $c$ with probability
$n_c(n_c-1)/(n(n-1))$.
The score is
\[
s = 1 - \frac{\sum_c W(S_c)}
{\mathbb{E}\left[\sum_c W(S_c)\right]},
\]
which equals $1$ for a perfect partition (each cluster internally
identical), is approximately $0$ for a random partition, negative for
an anti-clustered one, and exactly $0$ for the degenerate $k = 1$
``partition''---so a non-separating wrapper can never clear a
positive threshold.
The chance correction removes the mechanical dependence a raw
within/total ratio has on $k$ and on the cluster-size profile, so a
single threshold means the same thing everywhere in the tree.
This score is \emph{not} a Dasgupta cost delta and carries no
Dasgupta optimality interpretation; Dasgupta's cost
\parencite{dasgupta2016cost} is retained only as the whole-tree
Total Dasgupta Cost evaluation metric (Appendix~\ref{app:metrics}).

A split is accepted iff
$s \geq \epsilon_{\text{sep}}$ (the
\texttt{separationEpsilon} parameter) and every child satisfies
$|S_c| \geq n_{\min}$, with the parent node's recursion depth
hard-capped at $D_{\max}$.
The threshold has no closed-form optimum and trades over-splitting
against under-splitting; its canonical value is
$\epsilon_{\text{sep}} = 0.01$, subject to final calibration.
% TODO(calibration): finalize once DAG construction is frozen
The clustering proposes a small number of children ($k \geq 2$),
and $D_{\max} = 8$ caps recursion depth.
The greedy acceptance rule provides no worst-case global guarantee;
the \emph{same} score at the \emph{same}
$\epsilon_{\text{sep}}$ also gates sibling merging, sibling
distinctness, and residual-split viability
(Section~\ref{sec:meth-topology}), unifying these decisions under a
single dimensionless separability scale.

\section{Polyhierarchy and Cross-Links}\label{subsec:polyhierarchy}

The DAG structure---as distinct from a tree---arises exclusively from
topology operations in Phase~5; construction-time routing is weighted
multi-membership (Section~\ref{sec:meth-routing}), with the
\textit{primary} leaf defined as the argmax of the normalised
membership weights---a bookkeeping convention for aggregation, not an
exclusivity property of the fits.

Node-level polyhierarchy is introduced by two gated bridge sources,
both of which operate on the same admission rule as routing: a leaf
$L'$ qualifies as a secondary membership destination for a query iff
its posterior responsibility clears the \texttt{membershipFloor}
along the routing walk.
There is no separate concentration-scaled margin; the floor is a
probability and means the same thing everywhere in the tree.
(At arena time, qualifying leaves are additionally capped at
\texttt{maxLeafAssignments} per query as a judge-call-cost bound.)

\textbf{Source-A: residual bridges.}
Residual nodes (Section~\ref{sec:arch-dag}) that receive no primary
queries after routing are nonetheless retained as bridge targets: any
query whose secondary membership at a residual node clears
the floor is linked to that node via a secondary
edge, giving the residual node evaluable membership without altering
any query's primary leaf.

\textbf{Source-B: cross-domain bridges.}
After each splitting phase, \texttt{TaxonomyMerger} evaluates
candidate secondary edges between a query's primary leaf $L$ and every
other leaf $L'$ (including leaves under a different anchor) whose
secondary membership for $q$ clears the floor.
A candidate edge is inserted only if it passes four sequential gates:
deduplication against existing edges; a coherence gate, thresholded
at \texttt{fusionSimilarityThreshold} (canonical $0.90$), rejecting
edges where the resulting secondary membership would degrade $L'$'s
internal coherence below that value; a distinctness gate, on the same
\texttt{fusionSimilarityThreshold} scale, requiring the
primary and secondary leaf profiles to remain
cosine-distinct; and a cycle check combined with a cross-domain
dominant-category gate that blocks bridges which would make a leaf's
secondary membership dominate its primary-anchor identity.
A directed edge is added from the lowest common ancestor of $L$ and
$L'$ to $L'$ when all gates pass, giving $L'$ a second parent in the
DAG while $L$ retains its terminal (leaf) status and its status as
$q$'s primary leaf.
This preserves the invariant that a leaf node has no outgoing
parent-child edges.
A transitive reduction sweep follows both sources to maintain the
minimal DAG.

Evaluation-time multi-leaf assignment of held-out queries---the same
membership-floor trickle, applied read-only to the frozen snapshot
and capped at \texttt{maxLeafAssignments}---is
distinct from this node-level polyhierarchy: it does not modify the
frozen DAG topology but determines which leaves a given query
contributes comparisons to during the arena run
(Section~\ref{sec:arch-arena}).

\section{Trickle Routing and Polyhierarchy}\label{sec:meth-routing}

At node $v$ with children $\{c_1, \ldots, c_K\}$, a query embedding
$\mathbf{x}$ is scored against each child by that child's own vMF
log-density,
\[
f_k(\mathbf{x})
= \log C_d(\hat{\kappa}_{c_k})
+ \hat{\kappa}_{c_k} \hat{\boldsymbol{\mu}}_{c_k}^\top \mathbf{x},
\]
and the routing distribution is the temperature-1 softmax of these
log-densities,
\[
r_k(\mathbf{x})
= \frac{\exp f_k(\mathbf{x})}{\sum_{j=1}^{K} \exp f_j(\mathbf{x})},
\]
computed in log-space with log-sum-exp for multi-path aggregation.
This is the true Bayesian mixture posterior responsibility under
equal component priors: each child's fitted $\hat{\kappa}$ already
determines how sharp or soft its boundary is, so no separate
temperature or concentration-scaling knob is applied on top of it.

A single probability-scale parameter governs three decisions.
A child is admitted as a genuine membership destination iff
$r_k \geq \texttt{membershipFloor}$ (canonical value $0.10$, subject
to final calibration).
% TODO(calibration): finalize once DAG construction is frozen
Per-level transition probabilities are renormalised over the admitted
set, and the walk accumulates each path's log-probability from the
root; the \emph{same} floor bounds the cumulative path probability,
so a branch is not descended once its path probability decays below
the floor.
Finally, if \emph{no} child clears the floor, the position is
residual by construction: when \texttt{enableResidualRouting} is set
and the node is at depth $\geq 2$, the event is recorded as a
residual hit and feeds the Source-A bridges of
Section~\ref{sec:meth-topology}.
Terminology: a \textit{residual query} is a query retained at an
internal position because no child cleared the floor for it; a
\textit{residual node} (Section~\ref{sec:arch-dag}) is a leaf that
holds no primary queries---the two are related but distinct, and
Source-A bridges connect the former to the latter.

Membership weights over the reached leaves are normalised per query,
and the \textit{primary} leaf is the argmax of these weights---a
bookkeeping convention used for aggregation under the primary-path
convention, not an exclusivity property of the fits, which use the
full normalised weights.
Held-out queries at evaluation time are routed by exactly the same
distribution, applied read-only to the frozen snapshot, with the
reached-leaf set additionally capped at \texttt{maxLeafAssignments}
per query as a judge-call-cost bound; there is no separate
evaluation-time scoring model.
MMLU-Pro labels are never used in the routing distribution; they
are inspected only afterward, for the canonical-versus-adapted
divergence metrics (Section~\ref{sec:meth-validation}).

\section{Topological Optimisation}\label{sec:meth-topology}

A sequence of anchor-exempt operations runs after each splitting
phase.
\textbf{Sibling merge} pools two nodes when the chance-corrected
separation score of the two-cluster partition they form falls below
$\epsilon_{\text{sep}}$ (the same \texttt{separationEpsilon} and the
same score used by the splitting gate, so split and merge cannot
disagree about what ``separated'' means).
Historically this gate compared a $\kappa$-scaled symmetrised-KL
surrogate---often labelled a ``vMF JS divergence,'' although no
closed form for the vMF Jensen--Shannon divergence exists---whose
magnitude grew with $\kappa$ and became unsatisfiable exactly where
merge decisions mattered most; that surrogate no longer gates
anything.
\textbf{Starved-leaf pruning} removes a physical leaf whose total
branch query count falls far below the mean sibling size,
reassigning its members to the
highest-likelihood sibling.
% TODO(calibration): finalize once DAG construction is frozen
\textbf{Redundant single-child removal} eliminates a parent's sole
tree child: such a node separates no pair of queries its parent does
not already separate, so its marginal contribution to the hierarchy
objective is exactly zero, and it is removed unconditionally on
structural grounds.
% TODO(calibration): finalize once DAG construction is frozen
\textbf{Soft-membership refitting} (\texttt{refitMuPerIteration},
enabled under \texttt{DAG\_MAX} mode) recomputes each node's vMF
centroid from its full posterior-weighted membership rather than hard
assignment alone, subject to a no-update band: once a node's fresh
refit direction agrees with the previous one to within cosine
$0.9975$ the centroid is left unchanged for
that iteration, preventing noisy refitting once a node's membership
has effectively stabilised.
\textbf{Bridge insertion} admits polyhierarchical edges from two
gated sources, both operating on the membership floor of
Section~\ref{sec:meth-routing}.

\textit{Source-A (residual bridges).}
Residual leaves that received no primary queries after routing are
retained as bridge targets: any query whose secondary membership at
a residual node clears the floor is
linked to that node via a secondary edge, giving the residual node
evaluable membership without altering the query's primary leaf.

\textit{Source-B (cross-domain bridges).}
For a query whose primary leaf is $L$, a candidate secondary edge to
another leaf $L'$ (possibly under a different anchor) is proposed
whenever $L'$'s secondary membership for the query clears the
floor.
The candidate is accepted only if it passes four sequential gates:
deduplication against existing edges; a coherence gate rejecting
edges that would push $L'$'s internal coherence below the
\texttt{fusionSimilarityThreshold} (canonical $0.90$); a distinctness
gate, on the same \texttt{fusionSimilarityThreshold} scale, requiring
the primary and secondary leaf profiles to remain
cosine-distinct; and a combined cycle
check and cross-domain dominant-category gate that blocks bridges
which would make a leaf's secondary membership dominate its
primary-anchor identity.
When all gates pass, a directed edge is added from the lowest common
ancestor of $L$ and $L'$ to $L'$, giving $L'$ a second parent while
$L$ retains both its terminal (leaf) status and its status as the
query's primary leaf.

A transitive reduction sweep follows both bridge sources to maintain
the minimal DAG invariant.
