\chapter{Taxonomy Tuning Protocol and Ablations}\label{app:tuning-protocol}

This appendix details the mechanics behind the canonical hyperparameters
and construction ablations referenced in
Chapter~\ref{ch:experimental-design}; Table~\ref{tab:hyperparams}
lists the selected values.

\begin{table}[h]
\centering
\caption{Canonical taxonomy construction hyperparameters.}
\label{tab:hyperparams}
\begin{tabular}{lll}
\toprule
Parameter & Symbol/Name & Value \\
\midrule
Embedding dimension & $d$ & 256 (fixed) \\
Maximum recursion depth & $D_{\max}$ & 8 \\
Minimum cluster size & $n_{\min}$ & 50 \\
Unified separation threshold & $\epsilon_{\text{sep}}$ (\texttt{separationEpsilon}) & 0.01 \\
Membership floor\textsuperscript{a} & \texttt{membershipFloor} & 0.10 \\
Max leaf assignments (arena-time cap) & \texttt{maxLeafAssignments} & 5 \\
Fusion threshold & \texttt{fusionSimilarityThreshold} & 0.90 \\
Iteration budget & \texttt{numIterations} & 35 \\
DAG mode & \texttt{dagMode} & \texttt{DAG\_MAX} \\
\bottomrule
\end{tabular}
\par\smallskip
\footnotesize \textsuperscript{a}One parameter, three roles: child
admission (posterior responsibility floor), cumulative-path
probability bound, and residual-routing trigger.
Values are current canonical settings; final values
follow the calibration study conducted after the DAG-construction
pipeline is frozen.
% TODO(calibration): finalize once DAG construction is frozen
\end{table}

\section{Hyperparameter Selection Protocol}\label{sec:exp-tuning-protocol}

The L9($3^4$) orthogonal-array sweep documented here is
\textbf{historical}: it was conducted on a predecessor
parameterisation of the routing layer, and three of its four
factors---\texttt{routingSoftmaxTau}, \texttt{assignmentCosineGap},
and \texttt{deltaAssign}---were subsequently removed when routing was
unified under the single \texttt{membershipFloor} parameter
(Appendix~\ref{app:dag-formalism}).
Its protocol and gates are retained because they informed the
surviving canonical values (notably \texttt{minClusterSize}) and
because the selection machinery will be reused for the tuning study
of the current parameter set.
% TODO(calibration): finalize once DAG construction is frozen —
% document the membershipFloor/separationEpsilon tuning study here.
The design covered four factors---\texttt{minClusterSize},
\texttt{routingSoftmaxTau}, \texttt{assignmentCosineGap}, and
\texttt{deltaAssign}---each tested at three levels, run against a
Pareto selection procedure (\texttt{tools/tuning/pareto.py}) driven by
the harness \texttt{tools/tuning/tuning\_harness.py}.
The L9 levels were \texttt{minClusterSize} $\in \{50, 75, 100\}$,
\texttt{routingSoftmaxTau} $\in \{1.0, 1.5, 2.0\}$,
\texttt{assignmentCosineGap} $\in \{0.10, 0.15, 0.20\}$, and
\texttt{deltaAssign} $\in \{0.5, 1.0, 2.0\}$.

Configuration selection proceeds in three steps. Every candidate is
first checked against the eight hard gates in
Table~\ref{tab:pareto-gates}; a single failure excludes it. The same
candidate is then scored against the six soft gates in
Table~\ref{tab:pareto-gates}, whose outcome is recorded per candidate
but does not exclude it. All hard-gate survivors are then ranked by
Pareto dominance count over twelve coherence and calibration
objectives---five maximised (Weighted Leaf Purity, Dendrogram Purity,
Spherical Silhouette, total migration volume $m_{\text{total}}$ (the
fraction of queries whose anchor assignment changed relative to the
canonical partition; \emph{not} the RQ2 fidelity metric $\Delta\rho$,
which no tuning gate ever observes),
canonical--adapted Jaccard) and seven minimised (Total Dasgupta Cost,
Routing ECE, Brier score, no-match rate, small-leaf fraction,
$\kappa$-shrinkage mean, selected-node starved-leaf
fraction)---with raw Top-1 accuracy as the tiebreaker. Small-leaf
fraction, starved-leaf fraction, and canonical--adapted Jaccard
appear in both layers: the gate sets a hard floor, and the dominance
ranking additionally credits margin beyond it.

\begin{table}[h]
\centering
\caption{Configuration-selection gates used by the L9/Pareto tuning protocol.}
\label{tab:pareto-gates}
\begin{tabular}{p{5.4cm}p{3.6cm}}
\toprule
\multicolumn{2}{l}{\textit{Hard gates --- any failure excludes the candidate}} \\
\midrule
Acyclic & \texttt{true} \\
Duplicate bridges & $0$ \\
Orphaned nodes & $0$ \\
Assignment-cap rate & $\leq 0.20$ \\
Source-B bridge at depth $\geq 2$ & present \\
Small-leaf fraction & $\leq 0.50$ \\
Starved-leaf fraction (selected node) & $\leq 0.20$ \\
Source-B bridges per anchor (mean) & $\leq 3.0$ \\
\midrule
\multicolumn{2}{l}{\textit{Soft gates --- recorded, not excluding}} \\
\midrule
Average match count & $[1.05,\,1.30]$ \\
Top-1 routing accuracy & $\geq 75.6\%$ \\
Routing ECE & $\leq 0.25$ \\
Borderline rate & $[0.20,\,0.35]$ \\
Cross-anchor migration rate & $[0.10,\,0.30]$ \\
Canonical--adapted Jaccard & $[0.40,\,0.70]$ \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[h]
\centering
\caption{Parameters locked across the L9 sweep and ablations.}
\label{tab:tab-locked}
\begin{tabular}{ll}
\toprule
Locked parameter & Value \\
\midrule
$D_{\max}$ & 8 \\
\texttt{effectiveSupportFloor} & 2.0 \\
\texttt{fusionSimilarityThreshold} & 0.90 \\
\texttt{defaultKappaPrior} & 10.0 \\
\texttt{maxLeafAssignments} & 5 \\
\texttt{dagMode} & \texttt{DAG\_MAX} \\
\texttt{numIterations} & 35 \\
$d$ & 256 (except A1) \\
\bottomrule
\end{tabular}
\end{table}

The dynamic-temperature exponent $\gamma$ was \textbf{not} part of the
L9 design: because $\tau_{\text{eff}} = \tau\kappa^{\gamma}$ coupled
non-linearly to $\kappa$, it was ablated as a standalone factor at
$\{0, 0.5, 1.0\}$ (A11) rather than orthogonally; like the other
temperature knobs, $\gamma$ has since been removed along with the
mechanism it controlled, so A11 is historical. No reserved-arena-set
label, accuracy, or judge verdict is used by either the hard or soft
gates at any point in the L9/Pareto procedure; this is the mechanism
by which the tuning/test-leakage boundary is enforced in practice, not
merely asserted.

\section{Ablation Matrix}\label{sec:appendix-ablations}

Table~\ref{tab:ablations} summarises the ablation experiments run
against the canonical configuration, each varying exactly one
hyperparameter or structural choice while holding all others fixed.

\begin{table}[h]
\centering
\caption{Taxonomy construction ablation matrix.}
\label{tab:ablations}
\begin{tabular}{p{1.0cm}p{4.2cm}p{7.2cm}}
\toprule
ID & Factor & Levels tested \\
\midrule
A1 & Embedding dimension $d$ & $\{128, 256, 512\}$; canonical $= 256$ \\
A2 & $\kappa$ estimator corrections (Newton--Raphson refinement, Hornik--Gr\"un shrinkage, parent-anchored EB blend) & On (canonical) vs. off \\
A3$^\dagger$ & NMI variant & Overlapping vs. disjoint Shannon entropy \\
A4$^\dagger$ & Purity formula & DAG-dendrogram purity vs. standard tree purity \\
A5 & Separation threshold $\epsilon_{\text{sep}}$ & $\{0.01, 0.05, 0.10, 0.30\}$ \\
A6$^\ddagger$ & \texttt{deltaAssign} $\times$ \texttt{assignmentCosineGap} (joint; historical) & $\{0.5,1.0,2.0\} \times \{0.10,0.15,0.20\}$ \\
A7 & Iteration count & $\{10, 25, 50\}$ \\
A8 & Query set scope & Reserved (arena-excluded) vs. full corpus \\
A9 & Soft-membership refitting \texttt{refitMuPerIteration} & On (canonical) vs. off \\
A10 & Bridging / Source-A-B \texttt{enableBridging} & On (canonical) vs. off; Source-A only vs. Source-B only vs. both \\
A11$^\ddagger$ & Dynamic-temperature exponent $\gamma$ (historical) & $\{0.0, 0.5, 1.0\}$; canonical $= 0.0$ \\
\bottomrule
\end{tabular}
\end{table}

Each ablation is run across all three seeds and reported as mean
$\pm$ std following the statistical protocol of
Section~\ref{sec:exp-metrics}. A3 and A4 (marked $\dagger$) are not
system ablations in the same sense as the others: they vary the
\textit{metric computation formula} used to score a fixed,
already-constructed taxonomy, rather than a system parameter that
changes the taxonomy itself. They are retained as
\textit{reporting-sensitivity analyses}, testing whether headline
quality conclusions are an artefact of a specific metric definition.
A6 and A11 (marked $\ddagger$) are \textit{historical}: they ablated
the two nat-margin secondary-assignment gates and the dynamic routing
temperature of the predecessor routing parameterisation, all of which
were removed when routing was unified under \texttt{membershipFloor};
their rows are retained for provenance, and an equivalent ablation of
the current parameter set is pending.
% TODO(calibration): finalize once DAG construction is frozen
A9 and A10 isolate the \texttt{DAG\_MAX}-specific mechanisms
(soft-membership refitting and two-source bridging) whose marginal
contribution to taxonomy quality
and bridge coverage needs to be demonstrated, not merely described.

\section{Optional Secondary Diagnostics}\label{sec:appendix-optional-diagnostics}

The following conditions are not part of the minimum defensible
package (Table~\ref{tab:arena-conditions}) and are reported only
where run, as secondary diagnostics.

\subsection{Random Scheduler Ablation (C4)}

Because the scheduler is not a headline contribution
(\S\ref{sec:exp-rqmap}), the random scheduler ablation (C4) replaces
the composite-utility pair and query selection of
Section~\ref{sec:meth-scheduling} with uniformly random pair and
query selection, retaining the same per-pair and per-leaf budgets and
the same three-level stopping policy. It is compared against the
adaptive scheduler at matched final judge-call counts on total match
count, achieved Kendall $\tau$/pairwise-winner accuracy against
MMLU-Pro, and rounds to convergence. The budget-fidelity curve
($\rho$ or $\tau$ against cumulative judge calls) and a Wilcoxon
signed-rank test on per-leaf Kendall $\tau$ (chosen over a paired
$t$-test since per-leaf rank-agreement values are bounded and not
assumed normally distributed) compare the adaptive scheduler against
this ablation at matched judge-call budgets. An exhaustive
round-robin condition is not included: it is computationally
prohibitive at the 12-model arena scope and even less justified given
the scheduler is not a headline contribution.

\subsection{Verbosity Bias Probe}

Fifty judge matches are sampled uniformly from the canonical arena
run and re-judged with response position swapped, independent of the
standard dual-call protocol's own position randomisation, producing
an out-of-protocol audit sample. The position-flip rate (fraction of
re-judged matches whose verdict changes under the swap) and a
per-model verbosity delta $\Delta_{\text{verbosity}}$ (win-rate
difference attributable to response length after controlling for BT
score) are reported. This probe targets an anomaly observed in
preliminary runs (\texttt{arenabenchmark.log}) in which one evaluated
model's verdicts appeared to correlate with response length
independent of content quality; it is a targeted diagnostic, not a
general-purpose bias correction.
