\chapter{Introduction}\label{ch:introduction}

\section{Motivation}\label{sec:motivation}

\subsection{The Agentic Turn}\label{subsec:the-agentic-turn}

The field of Artificial Intelligence is undergoing a structural shift
from systems that respond to individual prompts towards systems that
pursue goals autonomously across extended sequences of
decisions~\parencite{wissuchek2025agentic}.
This transition, referred to in recent literature as the
\textit{agentic turn}, marks the evolution of Large Language Models
(LLMs) from passive text-generation interfaces into active
participants within larger software pipelines~\parencite{haidemariam2025turn}.
In agentic architectures, complex tasks are decomposed into sub-tasks
that are dispatched to specialised tools, APIs, or
models~\parencite{schick2023toolformer, yao2023react}.
The quality of this decomposition and dispatch depends critically on
one question: \textit{which model should handle a given sub-task?}

Answering this question requires a fine-grained, up-to-date
capability map of the available models.
Deployed agentic systems are now widespread in enterprise
contexts~\parencite{pwc2025survey}, yet the evaluation infrastructure
needed to build such a capability map remains largely absent.
Current benchmark leaderboards report a single aggregate score per
model, which conflates performance across semantically heterogeneous
tasks and provides no signal about domain-specific strengths or
failure modes.
A routing decision based on a global score is, at best, a coarse
approximation to the question the orchestrating agent actually needs
answered.

\subsection{The Evaluation Gap}\label{subsec:the-evaluation-gap}

The rapid deployment of LLMs and agentic systems has outpaced the
development of evaluation frameworks capable of distinguishing their
capabilities at the resolution required for informed
routing~\parencite{yan2025evaluation, shukla2025evaluating}.
Two structural limitations characterise current practice.

\subsubsection*{The Static Benchmark Problem}

Traditional benchmarks such as MMLU~\parencite{hendrycks2020measuring}
and its successor MMLU-Pro~\parencite{wang2024mmlu} assess capability
via fixed, editorially curated sets of questions organised into
predefined subject categories.
This design is inherently brittle along two axes.
First, as models improve, the hardest questions saturate:
top-performing models converge toward indistinguishable scores, and
the benchmark loses its ability to discriminate between closely
performing systems~\parencite{simmering2025benchmarks}.
Second, the category structure is fixed by human editors at
construction time.
Emerging capability dimensions---interdisciplinary topics, novel
application domains, or sub-skills that lie at the boundary between
existing categories---have no representation in the benchmark unless
the corpus is manually rebuilt, creating systematic blind spots.

\subsubsection*{The Scalar Fallacy and Aggregation Bias}

Even when evaluation signals are available, current ranking systems
reduce multidimensional capability profiles to a single scalar score,
such as an Elo rating or a weighted accuracy~\parencite{boubdir2024elo}.
This \textit{Scalar Fallacy} produces \textit{Aggregation Bias}:
averaging performance across all domains masks catastrophic failures
in specific subgroups and obscures capability profiles that differ
only in composition, not in mean~\parencite{shukla2025evaluating}.
Two models with identical global scores may differ sharply in their
domain profiles: one strong in formal reasoning and weak in applied
sciences, the other the reverse.
A global leaderboard assigns them identical routing priority for every
task, when a routing policy should assign them to entirely different
sub-tasks.

\subsection{The Limits of Conventional
Judging}\label{subsec:judging-limits}

Measuring performance on knowledge-intensive tasks requires a judge.
Expert human annotation remains the gold standard, but is
economically prohibitive at the scale required for continuous,
multi-model, multi-domain evaluation~\parencite{li2026agencybench}.
Automated scalar scoring introduces systematic calibration errors:
the judge has no stable internal reference for what a given score
means across domains, question types, or model generations.

Pairwise comparison resolves this calibration problem by reducing
judgement to a local, contextualised question: is answer~A better
than answer~B for this specific query~\parencite{zheng2024judging}?
Yet LLM-as-a-judge systems exhibit important biases even in pairwise
mode.
\textit{Positional bias} causes some judges to systematically favour
the answer presented first~\parencite{zheng2024judging}.
\textit{Verbosity bias} causes judges to favour longer answers
regardless of their actual quality~\parencite{li2024verbosity}.
Beyond observable biases, LLM judges also suffer from
\emph{guideline incompleteness}: any finite rubric under-specifies
the full evaluation space, leaving cases where the judge falls back
on undocumented internal heuristics.
Evaluating within semantically homogeneous groups of questions reduces
the verbosity--quality confound and enables tighter, domain-specific
rubrics that a general-purpose judge cannot provide.

\section{Problem Statement}\label{sec:problem-statement}

The preceding discussion identifies four concrete, interconnected
problems that remain unresolved in current evaluation practice.
Each is a necessary precondition for the next; problems~1 and~2
motivate the need for an induced evaluation structure, problem~3
establishes how that structure is made semantically coherent, and
problem~4 shows how coherent structure enables calibrated judging.
Contribution~1 addresses problems~1 and~3 by inducing evaluation
structure with geometrically coherent boundaries from a label-seeded
scaffold; Contribution~2 addresses problems~2 and~4 by producing
per-domain capability profiles via calibrated, domain-localised
pairwise judging.
An engineering component, the active match scheduler, additionally
reduces the judge-call cost implied by problem~4, but is not treated
as a headline research contribution (see Section~\ref{sec:research-goal}).

\begin{enumerate}

    \item \textbf{Fixed-corpus brittleness.}
    Existing benchmarks are organised around editorially defined subject
    categories that do not adapt to the evolving capability space of LLMs.
    A framework that uses editorial labels as a scaffold but refines
    domain structure geometrically can discover capability dimensions
    that matter without being permanently constrained by what the labels
    presuppose.

    \item \textbf{Scalar ranking and loss of capability resolution.}
    Aggregating performance into a single score conceals the domain
    structure of model capabilities.
    The routing problem in agentic systems requires per-domain rankings,
    not global leaderboards.
    A framework that produces capability profiles---vectors of scores
    indexed by semantic dimension---provides the resolution that scalar
    rankings discard.

    \item \textbf{Incoherent domain boundaries.}
    Induced evaluation structure is only well-defined if the resulting
    domains are semantically precise and internally homogeneous.
    A hierarchical decomposition of query space, derived from embedding
    geometry and formalised as a directed acyclic graph of semantic
    sub-domains, provides principled domain boundaries whose extent is
    defined by the data rather than by editorial judgement.

    \item \textbf{Judging at scale with controllable bias.}
    Automated evaluation must be scalable, yet current approaches trade
    off between calibration, bias, and domain sensitivity.
    A framework that combines pairwise comparison with bidirectional
    order evaluation, domain-localised judging, and principled
    statistical aggregation can control the known biases while remaining
    computationally tractable---but only if evaluation units are
    sufficiently homogeneous that outcomes are exchangeable within each
    domain.

\end{enumerate}

\section{Research Goal and Questions}\label{sec:research-goal}

\subsection*{Thesis Statement}

This thesis designs, implements, and evaluates \textbf{TaxoArena}: a
query-adaptive benchmark framework that tests whether answer-key-blind,
domain-conditioned pairwise LLM judges---evaluating models on held-out
queries routed through a geometrically induced semantic DAG---can
recover the per-domain capability rankings that MMLU-Pro produces
using ground-truth accuracy on selected high-support domain nodes,
and whether geometry-adapted domain groupings improve ranking
fidelity over the original editorial labels.

Two properties of the evaluation design are central to its validity.
The \textit{judge-generation phase} uses correct MMLU-Pro answers
from the construction query set to induce domain-specific evaluation
rubrics; the system is \textit{reference-informed} at construction
time.
The \textit{arena evaluation phase} evaluates held-out model
responses without access to those queries' correct answers; the
evaluation is \textit{answer-key-blind} at test time.
The MMLU-Pro accuracy oracle is revealed only after all arena matches
are complete.
This separation is formalised in Section~\ref{sec:info-separation}.

The title term \textit{Dynamic Domains} refers to this dual property:
domain boundaries are seeded from MMLU-Pro labels but subsequently
adapted by geometric evidence---queries may migrate to a
geometrically more appropriate domain, and new sub-domain leaves
emerge from recursive splitting---so that the evaluation dimensions
reflect the structure of the data rather than a fixed editorial
taxonomy.

\subsection*{Contributions}

\begin{enumerate}

    \item \textbf{Seed-guided geometric taxonomy construction.}
    TaxoArena induces a DAG taxonomy from query embeddings in a
    Matryoshka Representation Learning (MRL) space~\parencite{kusupati2022mrl},
    built over the full 14-domain MMLU-Pro corpus.
    The construction process begins from MMLU-Pro ground-truth domain
    labels, which seed the 14 top-level anchor nodes and provide a
    warm-start for the initial vMF distribution fits.
    From this starting point, the geometry takes over: queries may
    migrate to a more appropriate anchor based on embedding distance
    (producing an \textit{adapted partition} that may diverge from the
    original \textit{canonical} label-based partition), and each anchor
    is recursively split using a Dasgupta cost oracle~\parencite{dasgupta2016cost}
    until no further improvement is supported by the data.
    During construction, each query is assigned exclusively to one
    sibling node at each depth level, keeping the vMF and NiW
    distribution fits statistically clean.
    DAG structure---nodes with multiple parents---arises through
    topology operations (reparenting and cross-linking of
    structurally overlapping nodes) rather than through query
    multi-assignment during fitting.

    \item \textbf{Reference-informed, answer-key-blind arena evaluation.}
    For each leaf node, TaxoArena generates a domain-specific judge
    by presenting the construction queries---with their correct
    answers---to an LLM in a MapReduce induction pipeline.
    The resulting rubric encodes what correct reasoning looks like in
    that leaf's sub-domain, without referencing specific question
    answers.
    Given budget constraints, the arena roster (12 models) is evaluated
    on 1-2 ex-ante-selected, high-support domain nodes rather than across
    the full 14-domain taxonomy; the selection rule is specified in
    Chapter~\ref{ch:experimental-design}.
    Within each selected node, the arena evaluates held-out model
    responses pairwise using this rubric.
    Each comparison is run in both answer orders and the judge is asked
    to produce an analysis of each response before reaching a verdict
    (\textit{analysis-before-verdict prompting}); disagreements between
    the two orderings are treated as ties.
    Unlike the exclusive routing used during construction, arena routing
    permits a held-out query to enter multiple sufficiently compatible
    leaves when its embedding is close to the boundary between two
    sub-domains.
    Verdicts from multiple leaves are attributed fractionally (weight
    $1/k$ per leaf for a query entering $k$ leaves) when aggregating
    upward through the DAG, preventing double-counting.
    Results are aggregated via a Bradley--Terry model~\parencite{bradley1952rank,
        hunter2004mm}, producing coherent per-leaf, per-domain, and
    corpus-wide capability profiles.

\end{enumerate}

An active matchmaking scheduler that prioritises model pairs with
high rating uncertainty and declares per-leaf convergence from
Bradley--Terry standard errors is used to keep judge-call cost
tractable; it is an implementation detail supporting Contribution~2
rather than an independent contribution (Chapter~\ref{ch:methodology},
Section~\ref{ch:system-architecture}).

\subsection*{Research Questions}

\begin{enumerate}

    \item[\textbf{RQ1.}] Do answer-key-blind arena rankings agree with
    ground-truth MMLU-Pro accuracy on held-out queries, on the selected
    domain nodes?
    \textit{Primary metrics: Kendall} $\tau$\textit{, pairwise-winner
    accuracy, top-}$k$\textit{ Jaccard, and Spearman} $\rho$
    \textit{(secondary, reported with bootstrap confidence intervals
    given the small model count).}

    \item[\textbf{RQ2.}] Does evaluation within geometry-adapted leaf
    groupings improve ranking fidelity over flat MMLU-Pro domain
    groupings?
    \textit{Primary metric:} $\Delta\rho = \rho_{\text{adapted}} -
    \rho_{\text{canonical}}$, reported per selected node.

\end{enumerate}

As a secondary, non-headline diagnostic, Chapter~\ref{ch:results}
optionally reports whether the adaptive scheduler recovers the same
ranking at lower judge-call cost than an uninformed (random)
scheduling baseline, where this comparison is informative.

\section{Information Separation}\label{sec:info-separation}

A recurring concern in evaluation frameworks that use a corpus both
to construct their structure and to validate their output is
circularity: the same information that shaped the evaluation
instrument may also determine its apparent validity.
A closely related concern is specific to this thesis' hyperparameter
tuning: the L9 orthogonal design used to select DAG hyperparameters
(Chapter~\ref{ch:experimental-design}) must not be scored on the same
queries that later serve as the answer-key-blind arena test set, or
the "oracle revealed only post-hoc" property below would not hold.

TaxoArena addresses both concerns by partitioning the MMLU-Pro query
set into \textbf{three} disjoint subsets---construction (60\%),
tuning-validation (15\%), and final arena-test (25\%)---and enforcing
strict information boundaries across the phases of the pipeline.
DAG hyperparameters are selected using only the construction and
tuning-validation subsets; the arena-test subset is touched solely by
the arena-evaluation and validation-oracle phases.
Table~\ref{tab:info-separation-ch1} formalises these boundaries.

\begin{table}[h]
    \centering
    \caption{Information separation across pipeline phases.
    \checkmark~= observed; \texttimes~= not observed; (post) = observed
    only after all arena matches are complete.
    The answer-key-blind property holds for the Arena phase.}
    \label{tab:info-separation-ch1}
    \begin{tabularx}{\textwidth}{lcccc}
\toprule
\textbf{Phase} &
\textbf{Query text} &
\textbf{GT labels} &
\textbf{Correct answers} &
\textbf{Model responses} \\
\midrule
DAG construction &
\checkmark\ (60\%) & \checkmark\ (seed only) & \texttimes & \texttimes \\
\addlinespace
Judge generation &
\checkmark\ (60\%) & \texttimes & \checkmark\ (rubric induction) & \texttimes \\
\addlinespace
Hyperparameter tuning &
\checkmark\ (15\%) & \checkmark\ (validation only) & \texttimes & \texttimes \\
\addlinespace
Arena evaluation &
\checkmark\ (25\%) & \texttimes & \texttimes & \checkmark \\
\addlinespace
Validation oracle &
\checkmark\ (25\%) & \checkmark & \checkmark\ (post) & \checkmark\ (post) \\
\bottomrule
    \end{tabularx}
\end{table}

Four consequences of this separation are worth stating explicitly.

\textbf{Answer-key-blind arena.}
The arena judge applies the induced rubric to held-out queries whose
correct answers it has never seen.
The rubric encodes domain-specific quality criteria (what correct
reasoning looks like in this sub-domain) rather than specific answers
(what the correct choice is for a given question).
A judge that has learned what rigorous mathematical argumentation
looks like does not thereby know the answer to a specific unseen
integration problem.

\textbf{Construction--evaluation split.}
The DAG is built on the 60\% construction set; the 15\% tuning-validation
set is used exclusively to select DAG hyperparameters (Pareto gates
over the L9 design); the 25\% arena-test set is withheld until both the
taxonomy and its hyperparameters are frozen.
Reserved arena-test queries are then routed through the frozen DAG
using geometry alone (NiW posterior predictive scoring), with no label
information.
The trickle routing of reserved queries is therefore a genuine
out-of-sample test of the DAG's geometric structure, and the
agreement rate between geometric routing and MMLU-Pro labels---the
\textit{trickle test}---is a primary diagnostic of taxonomy quality.

\textbf{No tuning leakage.}
Because DAG hyperparameter selection uses only the construction and
tuning-validation subsets, no metric computed on the arena-test
subset ever feeds back into a system parameter.
This severs the leakage pathway that would otherwise arise if
tuning metrics (e.g. \texttt{Top1Accuracy}, \texttt{RoutingECE}) were
computed on the same queries later used to validate arena rankings.

\textbf{Oracle revealed post-hoc.}
MMLU-Pro per-domain accuracy scores for the evaluated models are
computed on the arena-test query set and compared to arena rankings
only after all matches are complete.
All hyperparameters are pre-committed before arena execution.
This prevents the arena-test set from functioning as a de facto
development set: the oracle cannot influence any system parameter.

\section{Thesis Structure}\label{sec:thesis-structure}

The remainder of this thesis is organised as follows.

Chapter~\ref{ch:literature-review} surveys five bodies of prior work:
LLM evaluation benchmarks, hierarchical taxonomy construction,
LLM-as-a-judge reliability, pairwise ranking models, and query
routing, with each section concluding with the design choice it
motivates in TaxoArena.
Chapter~\ref{ch:system-architecture} describes the system design
as three integrated pipelines: DAG construction (seed initialisation,
iterative splitting and topology optimisation, and convergence),
judge generation (MapReduce rubric induction per leaf), and arena
evaluation (routing, match scheduling, Bradley--Terry aggregation,
and upward propagation).
Chapter~\ref{ch:methodology} formalises the statistical and
geometric foundations: the vMF node model with bias-corrected MLE,
the Dasgupta cost oracle, the Bradley--Terry likelihood and MM
algorithm, the active scheduling utility function, and the
sufficient-statistic upward aggregation procedure.
Chapter~\ref{ch:experimental-design} specifies the experimental
configuration: corpus and domain selection, the three-way
construction/tuning-validation/arena-test split protocol with
pre-committed random seed, the L9 hyperparameter tuning protocol, the
ex-ante domain-node selection rule, DAG and arena hyperparameters, the
evaluated model set, judge configuration, baselines, and
pre-specified evaluation metrics and acceptance thresholds.
Chapter~\ref{ch:results} presents empirical results, covering DAG
construction quality, trickle-test domain agreement,
arena ranking validity~(RQ1), and adapted-vs-canonical
fidelity~(RQ2).
Chapter~\ref{ch:discussion} interprets the findings, addresses the
two research questions directly, and discusses construct validity,
the semantic-overlap interpretation of trickle-test divergence,
and limitations---including the selected-domain, proof-of-concept
scope of the arena evaluation.
Chapter~\ref{ch:conclusion} summarises the contributions and
outlines future work, including production router deployment,
multi-corpus transfer, and online domain discovery.