
\chapter{Formal Foundations}\label{ch:methodology}

This chapter formalises the statistical and algorithmic components
that TaxoArena's implementation actually executes: the directional
node model and its small-sample-corrected MLE, the fixed-dimension
Matryoshka embedding slice, the migration rule, the Dasgupta splitting
oracle, the dynamic-temperature trickle-routing softmax, the
Bradley--Terry estimator and its MM solver, the active scheduling
utility, the iteration-budget stopping policy, and upward score
aggregation.
Numerical-stability engineering that supports these computations but
is not itself a modelling choice---log-Bessel evaluation at high
dimension, the KDE bandwidth derivation, and the full Fisher
information matrix derivation---is summarised here and given in full
in Appendix~\ref{app:numerics}.

\section{Notation and Problem Formalisation}\label{sec:meth-notation}

Let $\mathcal{Q} = \{q_1, \ldots, q_N\}$ denote the corpus of
$N$ MMLU-Pro queries under consideration, each with a full-dimensional
Matryoshka embedding $\mathbf{x}_i \in \mathbb{R}^{4096}$ stored at unit
norm; the concrete value of $N$ and the domain scope are specified
in Chapter~\ref{ch:experimental-design}.
At depth $\ell$, the effective representation of $q_i$ is the prefix
slice $\mathbf{x}_i^{(\ell)} \in \mathcal{S}^{d-1}$ of fixed dimension
$d = 256$ (Section~\ref{sec:meth-mrl}).
Each query carries a ground-truth domain label
$y_i^{\text{MMLU}} \in \{D_1, \ldots, D_{14}\}$.

The corpus is partitioned once, before construction begins, into a
disjoint construction set $\mathcal{Q}_{\text{con}}$ and an evaluation
set $\mathcal{Q}_{\text{eval}}$, with $\mathcal{Q} = \mathcal{Q}_{\text{con}}
\cup \mathcal{Q}_{\text{eval}}$ and
$\mathcal{Q}_{\text{con}} \cap \mathcal{Q}_{\text{eval}} = \emptyset$.
$\mathcal{Q}_{\text{con}}$ is the only query set visible to the
taxonomy construction pipeline (anchor seeding, migration, splitting,
topological optimisation) and to reference-informed guideline
generation.
$\mathcal{Q}_{\text{eval}}$ is reserved exclusively for held-out
routing and arena evaluation and is never inspected during
construction; this boundary is a methodological constraint on the
experimental design, not a configuration choice, and is the basis of
the answer-key-blind claim formalised in
Table~\ref{tab:info-separation}.

The anchor set $\mathcal{A} = \{D_1, \ldots, D_{14}\}$ is in
one-to-one correspondence with the 14 MMLU-Pro domains; each anchor
is a depth-0 node from which the pipeline induces a sub-tree of
internal nodes and leaves $\mathcal{L}(D_k)$, with full leaf set
$\mathcal{L} = \bigcup_k \mathcal{L}(D_k)$.
The membership map $\hat{D}(q_i)$ assigns each query to the anchor of
its primary leaf after construction.

Two query populations are defined per anchor.
The \textit{canonical partition}
$Q_{D_k}^{\text{canonical}} = \{q_i : y_i^{\text{MMLU}} = D_k\}$ is
the MMLU-Pro label partition; the \textit{adapted partition}
$Q_{D_k}^{\text{adapted}} = \{q_i : \hat{D}(q_i) = D_k\}$ is the
partition induced by construction.
The two coincide absent cross-domain migration.

The primary objective is a per-anchor, per-model BT score
$s_{i,D_k}$ reflecting comparative response quality on
$Q_{D_k}^{\text{adapted}}$; the validation objective is the
rank-agreement between $\hat{\boldsymbol{s}}_{D_k}$ and the
ground-truth per-domain accuracy vector $\mathbf{A}_{D_k}$.

\section{Directional Node Model}\label{sec:meth-vmf}

\subsection{von Mises--Fisher Density and MLE}

The vMF distribution on $\mathcal{S}^{d-1}$ with mean direction
$\boldsymbol{\mu}$ and concentration $\kappa \geq 0$ has density
\[
p(\mathbf{x} \mid \boldsymbol{\mu}, \kappa)
= C_d(\kappa) \exp\!\left(\kappa \boldsymbol{\mu}^\top \mathbf{x}\right),
\qquad
C_d(\kappa) = \frac{\kappa^{d/2 - 1}}{(2\pi)^{d/2}\, I_{d/2 - 1}(\kappa)}.
\]
Given $N$ unit-vector observations, the MLE of $\boldsymbol{\mu}$ is
the normalised sample mean $\hat{\boldsymbol{\mu}} =
\bar{\mathbf{x}} / \|\bar{\mathbf{x}}\|_2$.
The Banerjee closed-form approximation~\parencite{banerjee2005vmf}
\[
\hat{\kappa}_{\text{ML}}
= \frac{\bar{R}\, d - \bar{R}^3}{1 - \bar{R}^2},
\qquad \bar{R} = \|\bar{\mathbf{x}}\|_2,
\]
avoids iterative root-finding for $\kappa$ and is accurate for
$d \geq 4$, $N \geq 5$, but is biased upward in high dimensions at
small sample sizes---an $O(d/N)$ bias that becomes severe when the
ratio $d/N$ is large.

Two safeguards against this bias are applied, both reflecting what the
implementation actually executes.
First, the closed-form estimate is multiplied by a small-sample
shrinkage factor
\[
\hat{\kappa}
= \hat{\kappa}_{\text{ML}} \cdot \frac{N - 1}{N + d - 2},
\]
a multiplicative correction in the spirit of the high-accuracy
$\kappa$ estimators of Hornik and Gr\"{u}n~\parencite{hornik2014vmf},
which pulls the closed-form estimate down toward its small-sample
expectation.
Second, the implementation flags any node whose $d/N$ ratio exceeds
$10$ as \textit{statistically unreliable}: when $d/N > 10$ the NiW
prior (Section~\ref{sec:meth-vmf}) is allowed to dominate the
posterior rather than the vMF MLE, and a warning is emitted.
This diagnostic threshold is the operational basis for the
joint choice of fixed dimension and minimum leaf size: at $d = 256$
and $n_{\min} = 50$ (Section~\ref{sec:exp-hyperparams}), accepted
leaves remain at $d/N \approx 5.1$, comfortably below the
implementation's unreliable-estimator regime, so the fixed-dimension
choice is made coherent with the minimum-cluster-size floor rather
than asserted independently.
Numerically stable evaluation of $\log I_\nu(\kappa)$ via the Debye
asymptotic expansion is given in Appendix~\ref{app:numerics}.

\subsection{NiW Auxiliary Regulariser}

A Normal-inverse-Wishart (NiW) posterior is fit to each node's member
embeddings to provide a tractable Bayesian membership-scoring model
for routing and cross-link insertion.
The NiW is \textit{not} the conjugate prior for the vMF---the vMF's
conjugate is the vMF itself~\parencite{gopal2014mises}---but is used
as an engineering regulariser whose posterior predictive is a
multivariate Student-$t$, accurate when $\kappa$ is moderate to large
so that the tangent-plane approximation to $\mathcal{S}^{d-1}$ holds.
At near-uniform nodes ($\kappa < 1$) the approximation degrades; such
nodes are flagged as geometric leaves and excluded from splitting.

The prior mean is the parent's vMF centroid, the prior pseudocount is
$\kappa_0 = 1.0$, the prior degrees of freedom are
$\nu_0 = d + 2$, and the prior scale is
$\boldsymbol{\Lambda}_0 = \lambda_0 \mathbf{I}_{d}$ with
$\lambda_0 = 1/(\hat{\kappa}_{\text{parent}} d)$.
Because the construction fixes $d = 256$ at every depth
(Section~\ref{sec:meth-mrl}), a diagonal approximation to
$\boldsymbol{\Lambda}_N$ is stored at all depths, reducing memory from
$O(d^2)$ to $O(d)$ per node and making cross-depth node comparisons
directly valid without down-projection.
The posterior update equations and the resulting Student-$t$
predictive density are given in Appendix~\ref{app:numerics}; the
predictive log-density is the scoring function used throughout
routing, migration, and cross-link decisions.
Node centroids are optionally smoothed across iterations by an
exponential moving average with weight \texttt{emaAlpha}
($\boldsymbol{\mu}^{(t)} = (1-\alpha)\,\hat{\boldsymbol{\mu}}^{(t)}
+ \alpha\,\boldsymbol{\mu}^{(t-1)}$), a numerical stabiliser on the
centroid refit distinct from the $\kappa$ shrinkage above.

\section{Fixed-Dimension Matryoshka Embedding Slice}\label{sec:meth-mrl}

MRL~\parencite{kusupati2022mrl} trains a single embedding model such
that every prefix slice $\mathbf{x}_{1:m}$ is itself a meaningful
embedding, allowing the full vector to be stored once and sliced at
query time.
Its key property is \textit{coarse-to-fine information packing}: the
most discriminative semantic directions are concentrated in the
leading coordinates, so any prefix $\mathbf{x}_{1:m'}$ is a coherent
lower-dimensional representation rather than a truncated artefact.

TaxoArena uses a single fixed slice
\[
d(\ell) = 256 \quad \text{for all depths } \ell,
\]
rather than a per-depth dimension schedule.
The choice balances two pressures.
A larger $d$ increases angular resolution and improves the
discriminability of fine-grained sub-topic splits, but the vMF
concentration estimate degrades as $d/N$ grows
(Section~\ref{sec:meth-vmf}); at $d = 256$ the implementation's
unreliability threshold ($d/N > 10$) is crossed only below $N \approx
26$, so the $n_{\min} = 50$ floor keeps every fitted leaf outside the
warning regime.
Fixing a single dimension additionally means all nodes---across all
depths---live on the same unit hypersphere, so sibling-merge,
passthrough-collapse, and bridge-coherence comparisons require no
cross-depth down-projection.
The coarse-to-fine property of MRL is exploited not by widening the
slice with depth but by the recursive structure itself: shallow nodes
separate coarse domain-level directions, while deeper splits resolve
progressively finer sub-topic structure within the same 256-dimensional
geometry.
Re-normalisation after slicing is mandatory, since MRL guarantees
semantic meaningfulness of $\mathbf{x}_{1:d}$ but not unit norm:
\(
\mathbf{x}^{(\ell)}
= \mathbf{x}_{1:d} / \|\mathbf{x}_{1:d}\|_2.
\)
The dimension itself is ablated (A1, Section~\ref{sec:exp-tuning-protocol})
at $\{128, 256, 512\}$ to confirm that $d = 256$ is a
quality-stable operating point rather than a fragile one.

\section{Seed Initialisation and Cross-Domain Migration}\label{sec:meth-migration}

At iteration 0, each anchor $D_k$ is initialised with the canonical
partition and a vMF/NiW self-model fit at $d = 256$; the anchor's
identity, depth, and \texttt{isAnchor} flag are set once and never
modified.

For each query $q_i$, the domain score under anchor $D_k$ is the
NiW log-posterior predictive
$z_{D_k}(\mathbf{x}_i) = \log p_{\text{NiW}}(\mathbf{x}_i \mid D_k)$.
Letting $D_{\text{seed}}(i)$ be $q_i$'s MMLU-Pro label and
$D^*(i) = \arg\max_k z_{D_k}(\mathbf{x}_i)$, query $q_i$ migrates from
$D_{\text{seed}}(i)$ to $D^*(i)$ iff
\[
z_{D^*(i)}(\mathbf{x}_i) - z_{D_{\text{seed}}(i)}(\mathbf{x}_i)
> \delta_{\text{mig}}.
\]
All $N$ migration decisions are evaluated in a single simultaneous
scoring pass before any reassignment, preventing cascading migrations
in which an early reassignment perturbs the NiW posteriors used to
score later queries.
Migration is applied only in the first construction iteration; a hard
cap of one pass prevents oscillation and stabilises
$Q_{D_k}^{\text{adapted}}$ before sub-domain splitting begins.
Anchor nodes are exempt from all destructive topology operations
regardless of migration volume, enforced by a
\texttt{TopologyViolationException} at the data-model level.

\section{Adaptive Splitting Oracle}\label{sec:meth-split}

Before applying vMF-EM refinement, a warm-start bisection is computed
via furthest-point $k$-means$++$ initialisation: the first centroid is
the node's vMF centroid, and the second is the member farthest by
angular distance, avoiding the dual-collapse failure mode of random
initialisation in high dimensions.

The Dasgupta cost of a hierarchical clustering
$T$ with cosine similarity $w_{ij} = \mathbf{x}_i^\top \mathbf{x}_j$
is $\text{cost}(T) = \sum_{i \neq j} w_{ij} |\text{leaves}(\text{LCA}_T(i,j))|$
\parencite{dasgupta2016cost}.
For a proposed bisection into $S_L, S_R$ with sizes $N_L, N_R$, the
normalised cost delta reduces to
\[
\delta_{\text{Dasgupta}}
= \frac{N_R W_L + N_L W_R}{N \cdot W_{\text{total}}},
\]
using the efficient identity
$W_L = \tfrac{1}{2}(\|\sum_{i \in S_L} \mathbf{x}_i\|_2^2 - N_L)$,
which reduces cost computation to $O(N_L d)$.
A split is accepted iff
$\delta_{\text{Dasgupta}} \geq \epsilon_{\text{sep}}$ (the
\texttt{separationEpsilon} parameter) and both children satisfy
$|S_L|, |S_R| \geq n_{\min}$, with the parent node's recursion depth
hard-capped at $D_{\max}$.
The threshold has no closed-form optimum and trades over-splitting
against under-splitting; at its canonical value
$\epsilon_{\text{sep}} = 0.01$ it functions primarily as a floor that
rejects only near-zero-cost (essentially arbitrary) splits, since
empirically observed deltas at accepted splits are an order of
magnitude larger.
The bisecting procedure always proposes exactly two children, and
$D_{\max} = 8$ caps recursion depth.
The greedy oracle provides no worst-case global-cost guarantee; the
same $\epsilon_{\text{sep}}$ quantity is reused as the sibling-merge
and passthrough-collapse threshold in Section~\ref{sec:meth-topology},
unifying three topology gates under a single separability parameter.

\section{Trickle Routing and Polyhierarchy}\label{sec:meth-routing}

At construction time, node $v$ with children $\{c_1, \ldots, c_K\}$
routes query embedding $\mathbf{x} \in \mathcal{Q}_{\text{con}}$ to
child $c_k$ with probability
\[
P_{\text{con}}(c_k \mid \mathbf{x}, v)
= \frac{\exp\!\left(\hat{\kappa}_{c_k} \hat{\boldsymbol{\mu}}_{c_k}^\top
\mathbf{x} / \tau_{\text{eff}}\right) p_{\text{NiW}}(\mathbf{x} \mid c_k)}
{\sum_{j=1}^{K} \exp\!\left(\hat{\kappa}_{c_j} \hat{\boldsymbol{\mu}}_{c_j}^\top
\mathbf{x} / \tau_{\text{eff}}\right) p_{\text{NiW}}(\mathbf{x} \mid c_j)},
\]
combining the vMF concentration term with the NiW posterior
predictive, computed in log-space with log-sum-exp for multi-path
aggregation.
The routing temperature is \textit{dynamic}: rather than a plain
$\tau$, the softmax divides by
\[
\tau_{\text{eff}} = \tau \cdot \kappa^{\gamma},
\]
where $\tau$ is the \texttt{routingSoftmaxTau} parameter,
$\kappa$ is the parent's mean child concentration, and
$\gamma \in [0,1]$ is the \texttt{tauKappaScalingFactor}.
At $\gamma = 0$ (the canonical value) the temperature reduces to the
plain $\tau$; larger $\gamma$ lets concentration dampen the softmax,
sharpening routing at high-$\kappa$ nodes.
$\gamma$ is ablated at $\{0, 0.5, 1.0\}$
(A11, Section~\ref{sec:exp-tuning-protocol}) as a standalone factor
separate from the L9 design.

At evaluation time, held-out queries
$\mathbf{x} \in \mathcal{Q}_{\text{eval}}$ are routed using the NiW
posterior predictive alone, with no
$\hat{\kappa}\hat{\boldsymbol{\mu}}^\top\mathbf{x}$ term:
\[
P_{\text{eval}}(c_k \mid \mathbf{x}, v)
= \frac{p_{\text{NiW}}(\mathbf{x} \mid c_k)}
{\sum_{j=1}^{K} p_{\text{NiW}}(\mathbf{x} \mid c_j)}.
\]
This asymmetry is deliberate: the vMF term encodes the frozen
snapshot's own directional geometry and is appropriate for the
queries that shaped that geometry, whereas the NiW posterior
predictive is a calibrated density that generalises to unseen queries
without re-fitting.
A depth-stratified residual gate retains a query at its current node
as a leaf when its top assignment probability under the applicable
routing distribution falls below a depth-dependent confidence floor:
the floor is $0.999$ at depths $0$--$1$ (where routing is expected to
be decisive) and $0.80$ at depth $\geq 2$ (where leaf-level assignment
is appropriately permissive), further scaled by the branching factor
$K$ so that the gate relaxes on highly branched nodes.
Residual routing is applied only when \texttt{enableResidualRouting}
is set and the node is at depth $\geq 2$, so the active residual gate
operates on depth-2-and-deeper nodes; at depths $0$--$1$ the high
$0.999$ floor is defined but residual retention is not invoked.
Queries that fail this gate are recorded as residual hits and feed
the Source-A bridges of Section~\ref{sec:meth-topology}.

Primary assignment---the single best-scoring child used for fitting
and for all downstream statistics under the primary-path
convention---remains exclusive at every depth: exactly one child is
designated primary per query per depth, regardless of how many
children clear the residual gate.
Secondary, multi-leaf assignment is governed by two nat-margin gates
on the accumulated log-probabilities of the routing walk.
The \textit{walk-breadth gate} \texttt{deltaAssign} controls how many
children the routing walk explores: a child is descended into only if
its log-probability is within \texttt{deltaAssign} nats of the
best sibling's.
The \textit{leaf-qualification gate} determines which reached leaves a
query is finally assigned to: a leaf $L'$ qualifies iff its accumulated
log-probability is within
\[
\text{margin} = \texttt{assignmentCosineGap} \cdot \kappa_{\text{eff}}
\]
nats of the best leaf's, where $\kappa_{\text{eff}}$ is the mean
concentration of the parent's children (coerced to $[1, 100]$).
Qualifying leaves are capped at \texttt{maxLeafAssignments} per query.
This gate directly scales the cosine gap by concentration---a query
must clear a larger log-probability margin to enter a second leaf when
the local geometry is sharply peaked ($\kappa$ large) than when it is
diffuse---and it is the sole basis for the bridges introduced in
Section~\ref{sec:meth-topology}.
Secondary assignment never overrides or duplicates the primary
assignment; it only creates candidate bridge evidence.
MMLU-Pro labels are never used in either routing distribution; they
are inspected only afterward, for the canonical-versus-adapted
divergence metrics (Section~\ref{sec:meth-validation}).

\section{Topological Optimisation}\label{sec:meth-topology}

A sequence of anchor-exempt operations runs after each splitting
phase.
\textbf{Sibling merge} pools two nodes when their vMF
Jensen--Shannon divergence falls below $\epsilon_{\text{sep}}$ (the
same \texttt{separationEpsilon} used by the splitting oracle).
\textbf{Starved-leaf pruning} removes a physical leaf whose total
branch query count falls below $0.2$ times the mean sibling size
(floored at $5$), reassigning its members to the
highest-likelihood sibling.
\textbf{Passthrough collapse} bypasses a single-child internal node
when its child's vMF JS-divergence falls below $\epsilon_{\text{sep}}$
\emph{and} its member count is below $n_{\min}$, promoting the child
up one level.
\textbf{Soft-membership refitting} (\texttt{refitMuPerIteration},
enabled under \texttt{DAG\_MAX} mode) recomputes each node's vMF
centroid from its full posterior-weighted membership rather than hard
assignment alone, subject to a no-update band: once a node's
dominant posterior mass exceeds $0.9975$ the centroid is frozen for
that iteration, preventing noisy refitting once a node's membership
has effectively stabilised.
\textbf{Bridge insertion} admits polyhierarchical edges from two
gated sources, both operating on the leaf-qualification margin of
Section~\ref{sec:meth-routing}.

\textit{Source-A (residual bridges).}
Residual leaves that received no primary queries after routing are
retained as bridge targets: any query whose secondary NiW posterior
score at a residual node clears the leaf-qualification gate is
linked to that node via a secondary edge, giving the residual node
evaluable membership without altering the query's primary leaf.

\textit{Source-B (cross-domain bridges).}
For a query whose primary leaf is $L$, a candidate secondary edge to
another leaf $L'$ (possibly under a different anchor) is proposed
whenever $L'$'s NiW posterior score for the query clears the
leaf-qualification gate.
The candidate is accepted only if it passes four sequential gates:
deduplication against existing edges; a coherence gate rejecting
edges that would push $L'$'s internal coherence below the
\texttt{fusionSimilarityThreshold} (default $0.92$); a distinctness
gate requiring the primary and secondary leaf profiles to remain at
least $0.92$ cosine-distinct; and a combined cycle
check and cross-domain dominant-category gate that blocks bridges
which would make a leaf's secondary membership dominate its
primary-anchor identity.
When all gates pass, a directed edge is added from the lowest common
ancestor of $L$ and $L'$ to $L'$, giving $L'$ a second parent while
$L$ retains both its terminal (leaf) status and its status as the
query's primary leaf.

A transitive reduction sweep follows both bridge sources to maintain
the minimal DAG invariant.

\section{Bradley--Terry Estimation}\label{sec:meth-bt}

\subsection{Model and MM Solver}

The BT model assigns each model $m_i$ a latent skill $s_i$ with
$P(m_i \succ m_j) = \sigma(s_i - s_j)$~\parencite{bradley1952rank}.
A zero-sum constraint $\sum_i s_i = 0$ fixes the additive degree of
freedom; ties contribute a half-win to each model, so
$W_{ij} + W_{ji} = N_{ij}$ without an explicit draw parameter.
The Minorisation-Maximisation update~\parencite{hunter2004mm}
\[
s_i^{(t+1)} = \log\!\left(
\frac{W_i}{\sum_{j \neq i} N_{ij} / \sigma(s_i^{(t)} - s_j^{(t)})}
\right),
\qquad W_i = \sum_{j \neq i} W_{ij},
\]
is applied cyclically with the zero-sum constraint re-enforced after
each sweep, until $\max_i |s_i^{(t+1)} - s_i^{(t)}| < 10^{-6}$ or a
500-iteration cap is reached.

\subsection{Standard Errors and Identifiability}

Standard errors are $\hat{\sigma}_i =
\sqrt{[\mathcal{I}(\hat{\mathbf{s}})^{-1}]_{ii}}$ from the observed
Fisher information matrix (full form in
Appendix~\ref{app:numerics}), using the pseudoinverse to account for
the rank-one deficiency induced by the zero-sum constraint.
BT parameters are identified only up to an additive constant per
connected component of the comparison graph; the connectivity
bootstrap phase of the scheduler (Section~\ref{sec:meth-scheduling})
guarantees every pair receives at least one comparison before
utility-driven scheduling begins.

\section{Active Match Scheduling}\label{sec:meth-scheduling}

The composite utility of comparison triple $(i,j,L)$ combines a
structural exploration term and an uncertainty resolution term under
a maturity-decaying mixing parameter $\alpha$:
\[
U(i,j,L) = \left[\alpha\, U_{\text{struct}}(i,j,L)
+ (1-\alpha)\, U_{\text{resolve}}(i,j,L)\right]
\times \left(1 - 0.3\, \frac{n_{ij}^{(L)}}{B_{ij}}\right).
\]
$U_{\text{struct}}$ rewards closely-ranked pairs
($\exp(-|\text{rank}_i - \text{rank}_j|)$).
$U_{\text{resolve}}$ is the binary entropy of the current win
probability $p_{ij} = \sigma(\hat{s}_i - \hat{s}_j)$, divided by the
combined variance $\hat{\sigma}_i^2 + \hat{\sigma}_j^2$, and weighted
by a KDE valley weight favouring midpoints in low-density regions of
the score distribution (bandwidth and weight derivation in
Appendix~\ref{app:numerics}).
The entropy term is maximised when $p_{ij} \approx 0.5$, prioritising
pairs whose current ranking is genuinely ambiguous; the variance
division then downweights pairs whose standard errors are already
large, since a single additional comparison contributes little
information relative to the pair's existing noise floor, and
upweights pairs that are ambiguous \textit{despite} already having
tight, well-estimated scores---the case where one more comparison is
most likely to be decisive.
This is the internally consistent reading of the utility term: entropy
identifies ambiguity, and dividing by variance targets that ambiguity
toward pairs where additional evidence is most informative rather than
toward pairs that are simply noisy.
The mixing parameter decays as
$\alpha = 0.20 + 0.80\,\exp(-2.5\,\text{maturity})$, where
\textit{maturity} is a weighted combination of standard-error
reduction and completed-pair fraction, so $\alpha$ falls from $1.0$
toward approximately $0.20$ as a leaf matures---exploration-dominated
early, uncertainty-resolution-dominated late.
This scheduler-internal $\alpha$ is distinct from the centroid
\texttt{emaAlpha} of Section~\ref{sec:meth-vmf}.
Within a scheduled pair, queries are selected by descending cosine
similarity to the leaf's vMF centroid, prioritising representative
queries.

\section{Stopping Policy}\label{sec:meth-stopping}

A pair $(i,j)$ at leaf $L$ resolves when any of three conditions
holds: separation
$|\hat{s}_i - \hat{s}_j| \geq 3.0(\hat{\sigma}_i + \hat{\sigma}_j)$;
budget exhaustion $n_{ij}^{(L)} \geq B_{ij}$; or utility floor
$U(i,j,L) < U_{\min}$.
A leaf converges when all pairs have resolved and the top-two
separation guard
$|\hat{s}_1^{(L)} - \hat{s}_2^{(L)}| > 1.5(\hat{\sigma}_1^{(L)} +
\hat{\sigma}_2^{(L)})$ holds, preventing premature convergence when
leading models remain statistically indistinguishable.
The arena halts globally when the query-weighted converged fraction
$f_{\text{conv}} = \sum_{L \in \mathcal{C}} |Q_L| /
\sum_{L \in \mathcal{L}} |Q_L|$ reaches $f^*$ for $R_{\text{stable}}$
consecutive rounds, with a hard escape valve at $B_{\max}$ total judge
calls.
The taxonomy construction loop, by contrast, runs for a fixed
iteration budget (\texttt{numIterations}, canonical $35$): there is no
structural early-stopping gate in construction, and the structural
edit distance between consecutive DAG snapshots is reported as a
convergence \textit{diagnostic} (quiescence indicator) rather than a
stopping criterion.

\section{Upward Aggregation to GT Domains}\label{sec:meth-aggregation}

Domain-level win and comparison counts pool leaf-level sufficient
statistics under the primary-path convention,
$W_{ij,D_k} = \sum_{L \in \mathcal{D}(D_k)} W_{ij}^{(L)}$ and
analogously for $N_{ij,D_k}$, and a fresh MM fit produces
$\hat{\mathbf{s}}_{D_k}$.
Leaf-level parameters are never averaged: BT identification is fixed
only up to a per-component additive constant, so averaging
differently-normalised leaf scales produces no interpretable result,
whereas pooling counts and refitting yields a single shared scale.
A fractional-weight sensitivity analysis, in which a leaf with $k$
primary parents contributes $(1/k)$ of its counts to each, is
recorded under a separate \texttt{aggregationConvention} field for
comparison in Chapter~\ref{ch:results}.

Two parallel tracks are maintained per anchor: \textit{Track A
(canonical)} groups verdicts by MMLU-Pro label, producing
$\rho_{\text{canonical}}$; \textit{Track B (adapted)} groups by the
adapted partition, producing $\rho_{\text{adapted}}$.
The migration effect is
$\Delta\rho = \rho_{\text{adapted}} - \rho_{\text{canonical}}$; a
positive value indicates that geometric re-partitioning improves
rank agreement with ground-truth accuracy.

\section{Validation Against MMLU-Pro}\label{sec:meth-validation}

Ground-truth per-domain accuracy
$A_{i,D_k} = |\{q \in Q_{D_k} : m_i(q) = y_q^{\text{MMLU}}\}| / |Q_{D_k}|$
is computed on both partitions.
The judge rubric is \textit{reference-informed}: construction-set
reference answers inform the domain-conditioned evaluation criteria
generated by \texttt{GuidelineGenerator} (Section~\ref{subsec:judge}),
so that the judge's notion of quality is shaped by what correct
responses look like in the domain.
Ground-truth accuracy itself, and the reference answers for
$\mathcal{Q}_{\text{eval}}$ specifically, are never inspected during
arena evaluation; $A_{i,D_k}$ is computed strictly after all verdicts
are recorded and is used only for external, post-hoc validation.
Four rank-agreement metrics are reported per domain: Spearman $\rho$;
Kendall $\tau$ (preferred when $M$ is small); pairwise-winner
accuracy; and top-$k$ Jaccard agreement for $k \in \{3,5\}$.
The budget-fidelity curve plots $\rho$ against cumulative judge calls
for the adaptive scheduler versus random and round-robin baselines,
quantifying the marginal value of adaptive scheduling and the budget
at which additional calls yield diminishing returns.

\section{Methodological Assumptions}\label{sec:meth-assumptions}

Six assumptions underlie the methodology, each paired with a
diagnostic or sensitivity check: directional geometry reflects
semantic compatibility (checked via spherical silhouette score);
single-component vMF adequacy per node (checked via the Dasgupta
split test); local BT transitivity within coherent domains (supported
by domain-conditioned judging); judge reliability under domain
conditioning (partially tested by calibration analysis, not
structurally guaranteed); MMLU-Pro domains as a useful reporting
vocabulary (quantified by $\Delta\rho$); and query reuse inducing
within-query dependence across comparisons, which the Fisher
information standard errors treat as independent and therefore
slightly underestimate.

\section{Chapter Summary}

This chapter formalised the vMF/NiW node model with its
small-sample-corrected MLE and the $d/N$-coherence rationale for the
fixed $d = 256$ slice, the migration rule with its single-pass
oscillation guard, the Dasgupta splitting oracle with its unified
$\epsilon_{\text{sep}}$ topology gates, the log-space trickle-routing
softmax with dynamic temperature $\tau_{\text{eff}} = \tau\kappa^{\gamma}$,
the depth-stratified residual gate and the two nat-margin
secondary-assignment gates, the topological optimisation pass with
soft-membership refitting and two-source bridge insertion, the
half-win BT estimator with its MM solver and Fisher-information
standard errors, the composite-utility active scheduler, the
three-level arena stopping policy alongside the fixed-iteration
construction budget, sufficient-statistic pooling for upward
aggregation under the primary-path and fractional-weight conventions,
and the four rank-agreement validation metrics against MMLU-Pro.
Numerical-stability procedures supporting these models---log-Bessel
evaluation, the NiW posterior update algebra, the KDE bandwidth
derivation, and the full Fisher information matrix---are given in
Appendix~\ref{app:numerics}.
Hyperparameter values are specified and justified in
Chapter~\ref{ch:experimental-design}.
